Contextual recruitment of selective attention can be updated via changes in task-relevance

### 2018-12-09 manuscript.Rmd
first draft

keywords          : "Attention, Memory, Cognitive Control, Conflict Adaptation, Context-specific, Task Relevance"
wordcount         : "5946"

@@

output:
 papaja::apa6_pdf
---

```{r setup, include=FALSE, eval=TRUE,echo=FALSE}
# options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(message=FALSE,warning=FALSE, cache = FALSE)

library("dplyr")
library("afex")
library("papaja")
```

#  Introduction

Selective attention processes are commonly investigated using interference paradigms such as the Stroop [-@stroop_studies_1935] and flanker [@eriksen_effects_1974] tasks, where participants identify a target while ignoring a response-congruent or -incongruent distractor. Performance is typically better on congruent versus incongruent trials and the difference-the congruency effect-taken as an index of attentional priorities. Large congruency effects are thought to reflect ineffective filtering of the distracting stimuli whereas small congruency effects are thought to reflect effective filtering. By probing factors that systematically alter congruency effects, we can then make inferences about processes that control attentional filtering. For example, manipulating the frequency of conflict via the proportion of congruent versus incongruent trials has shown to influence the size of the congruency effect. Typically, a high proportion congruent experiment produces large congruency effects, whereas a low proportion congruent experiment produces small congruency effects [@logan_when_1979; @lowe_selective_1982; @west_effects_1998]. This result is usually  explained as strategic control, where participants increase attentional control under high-conflict demands and relax attentional control under low-conflict demands [@logan_attention_1980; @logan_strategies_1984; @logan_when_1979; @lowe_selective_1982]. Recent work however, has demonstrated that attentional control is not only adjusted by top-down regulation, but can also be triggered automatically by environmental cues [@brosowsky_memory-guided_2018; @bugg_support_2012; @egner_creatures_2014; @fischer_predicting_2015; @king_priming_2012; @mayr_outsourcing_2007].

@@

To date, we are aware of only one study that has looked at task-relevance within the CSPC literature. Crump, Vaquero, & Milliken [-@crump_context-specific_2008] used a Stroop prime-probe task where the distractor word is presented first followed by a color patch probe. Critically, they used the shape of the color patch probe, either a square or a circle, as the contextual dimension; one associated with a high frequency of conflict, the other with a low frequency. In one experiment participants were made aware of the shape-proportion contingencies but given no additional instructions. In the second, participants were given a secondary task to count the number of trials that contained a square. The first experiment failed to find evidence for context-specificity, whereas the second did. Although not testing task-relevance directly, [-@canadas_social_2013]) found similar effects. In this study they used images of male and female faces as context cues in a flanker task. When given instructions to think of the faces as members of the gender categories they found CSPC effects, but when given instructions to think of the faces as individuals they did not. These results suggest that making a context dimension relevant to the on-going task is important for producing CSPC effects.

@@

Some prior suggests that the task-relevance of contextual information is important for establishing new associations between context cues and attentional priorities. Simply put, CSPC effects emerge when context cues are somehow made relevant to the task, but do not when made irrelevant [@canadas_social_2013; @crump_context-specific_2006; @crump_context-specific_2008]. However, it still remains unclear how task-relevance may or may not inform the use of previously learned associations. In the current study, we examined whether manipulating task-relevance would cause participants to ignore a previously learned association and adopt a new context-specific control strategy.

@@

The critical measure of interest is the congruency effect produced in the frequency unbiased context. If, on the one hand, task-relevance has no impact on the use of context-attention associations we would expect no differences between the congruency effects when the high-conflict context dimension is made task-relevant or the low-conflict context dimension is made task-relevant. This could occur because all three contexts are treated as individual contexts throughout the whole experiment [e.g., @canadas_social_2013]. Or it could occur because associations formed in the first block interfere with learning new associations in the second [e.g., @brosowsky_context-specific_2016]. On other hand, task-relevance might not only dictate how associations are formed, but also whether learned associations are used. In this case, we would expect that there will be smaller congruency effects when the high-conflict context dimension is made task-relevant as compared to when the low-conflict context dimension is made task-relevant. This would demonstrate that participants were able to ignore a previously learned association and adopt a new context-specific control strategy.

A secondary goal was to conceptually replicate previous findings and test the generalizability of any task-relevance effects. Therefore, we included three conditions that differed only in the kinds of stimuli used to create context cues. In one condition we used face images where context dimensions were defined by social categories, similar to Cañadas et al.,  [-@canadas_social_2013]. However, they found that social categorization occurred spontaneously and it was unclear what impact that would have on the current results. Therefore, we included another condition that used images of simple objects where context dimensions were defined by object features. Finally, Cañadas et al.[-@canadas_social_2013] also found that context effects generalized to novel face images. In our third condition we used non-repeating face exemplars to examine whether any task-relevance effects would also generalize in our design or if learning would be image-specific.

@@

```{r figure1, echo=FALSE, fig.cap="(ref:figure1)", fig.width=6.7,fig.height=4.51}
knitr::include_graphics("images/Task.jpg")
```

@@

All participants were recruited from Amazon Mechanical Turk (AMT) and compensated $2.00 for participating. The amount compensated was calculated by estimating the maximum amount of time required to complete each experiment and multiplying by $6.00 per hour. For each experiment the number of HITs (Human intelligence tasks, an Amazon term for a work-unit) refers to the number of participants who initiated the study. Participants were included in the study if they completed all trials. We posted 150 HITs and 145 participants completed all trials.

@@

@@

```{r analysis}
source("data-analysis/analysis.R")

```

```{r figure2, echo=FALSE, fig.cap="(ref:figure2)", fig.width=6.875, fig.height= 4}
figure2
```

@@

Similarly, the corresponding error analysis also resulted in a significant two-way interaction between the task-relevant context and unbiased-item congruency, `r RT_ANOVA$Task_Relevant_Context_Congruency`, and a non-significant three-way interaction between task-relevant context, unbiased-item congruency, and context-type, `r RT_ANOVA$Condition_Task_Relevant_Context_Congruency`. Therefore, the congruency effects in error rates were also smaller when the context dimension associated with high conflict was made task-relevant, corroborating the results of the reaction time analysis.

@@

The aim of the current study was to investigate whether task-relevance plays a role in the contextual recruitment of selective attention. Specifically, we tested whether manipulating the relative task-relevance of context cues could cause participants to ignore a previously learned context-association and apply a new association. We created a frequency unbiased context cue that shared features with two frequency biased contexts and included a feature-counting task to manipulate the task-relevance of context dimensions across two blocks of trials. Critically, halfway through the experiment participants received new instructions changing the task-relevant feature from one frequency biased cue to the other.

@@

In sum, our results provide new evidence that changes in task-relevance can update the contextual recruitment of selective attention in a CSPC flanker task. Importantly, we show that making one feature task-relevant produced a CSPC effect, even in the presence of a competing contextual cue. Moreover, changing the task-relevant context-cue midway through the experiment was accompanied by predictable changes in the congruency effects. This effect was found to be generalizable across two different kinds of stimuli and occurred even when using non-repeating images. These results implicate an important role for task-relevance in producing context-dependency and are consistent with memory-based accounts of CSPC phenomena.

@@

>>>>

keywords          : "Attention, Memory, Cognitive Control, Conflict Adaptation, Context-specific, Task Relevance"
wordcount         : "4402"

@@

#output: 
#  word_document

output:
 # papaja::apa6_word: default
  papaja::apa6_pdf: default
  #redoc::rdocx_reversible

knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding,
  output_format = "all") })

header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage[table]{xcolor}
  - \usepackage{wrapfig}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{setspace}\singlespacing 
  - \newenvironment{zeroindent}
     {\par\setlength{\parindent}{0pt}}
     {\par}  
  - \setlength{\parskip}{0.25cm plus4mm minus3mm}

---
```{r setup, include=FALSE, eval=TRUE,echo=FALSE}
library("dplyr")
library("afex")
library("papaja")
library("magrittr")
library("kableExtra")

#As of 2018/12/07 apa_print() requires development version of Papaja
#devtools::install_github("crsh/papaja")
library(papaja)
```

Selective attention is commonly investigated using interference paradigms such as the Stroop [-@stroop_studies_1935] and flanker [@eriksen_effects_1974] tasks, where participants identify a target while ignoring a response-congruent or -incongruent distractor. Performance is typically better on congruent versus incongruent trials and the difference-the congruency effect-taken as an index of attentional priorities. Large congruency effects are thought to reflect ineffective filtering of the distracting stimuli whereas small congruency effects are thought to reflect effective filtering. By probing factors that systematically alter congruency effects, we can then make inferences about processes that control attentional filtering. For example, manipulating the frequency of conflict via the proportion of congruent versus incongruent trials has shown to influence the size of the congruency effect. Typically, a high proportion congruent experiment produces large congruency effects, whereas a low proportion congruent experiment produces small congruency effects [@logan_when_1979; @lowe_selective_1982; @west_effects_1998]. This result is usually  explained as strategic control, where participants increase attentional control under high-conflict demands and relax attentional control under low-conflict demands [@logan_attention_1980; @logan_strategies_1984; @logan_when_1979; @lowe_selective_1982]. Recent work however, has demonstrated that attentional control is not only adjusted by top-down regulation, but can also be triggered automatically by environmental cues [@brosowsky_memory-guided_2018; @bugg_support_2012; @egner_creatures_2014; @fischer_predicting_2015; @king_priming_2012; @mayr_outsourcing_2007].

@@

To date, we are aware of only one study that has looked at task-relevance within the CSPC literature. Crump, Vaquero, & Milliken [-@crump_context-specific_2008] used a Stroop prime-probe task where the distractor word is presented first followed by a color patch probe. Critically, they used the shape of the color patch probe, either a square or a circle, as the contextual dimension; one associated with a high frequency of conflict, the other with a low frequency. In one experiment participants were made aware of the shape-proportion contingencies but given no additional instructions. In the second, participants were given a secondary task to count the number of trials that contained a square. The first experiment failed to find evidence for context-specificity, whereas the second did. Although not testing task-relevance directly, @canadas_social_2013] found similar effects. In this study they used images of male and female faces as context cues in a flanker task. When given instructions to think of the faces as members of the gender categories they found CSPC effects, but when given instructions to think of the faces as individuals they did not. These results suggest that making a context dimension relevant to the on-going task is important for producing CSPC effects.

@@

Prior work suggests that the task-relevance of contextual information is important for establishing new associations between context cues and attentional priorities. Simply put, CSPC effects emerge when context cues are somehow made relevant to the task, but do not when made irrelevant [@canadas_social_2013; @crump_context-specific_2006; @crump_context-specific_2008]. However, it still remains unclear how task-relevance may or may not inform the use of previously learned associations. In the current study, we examined whether manipulating task-relevance would cause participants to ignore a previously learned association and adopt a new context-specific control strategy.

@@

The critical measure of interest is the congruency effect produced in the frequency unbiased context. If, on the one hand, task-relevance has no impact on the use of context-attention associations we would expect no differences between the congruency effects when the high-conflict context dimension is made task-relevant or the low-conflict context dimension is made task-relevant. This could occur because all three contexts are treated as individual contexts throughout the whole experiment [e.g., @canadas_social_2013]. Or it could occur because associations formed in the first block interfere with learning new associations in the second [e.g., @brosowsky_context-specific_2016]. On other hand, task-relevance might not only dictate how associations are formed, but also whether learned associations are used. In this case, we would expect smaller congruency effects when the high-conflict context dimension is made task-relevant as compared to when the low-conflict context dimension is made task-relevant. This would demonstrate that participants were able to ignore a previously learned association and adopt a new context-specific control strategy.

A secondary goal was to conceptually replicate previous findings and test the generalizability of any task-relevance effects. Therefore, we included three conditions that differed only in the kinds of stimuli used to create context cues. In one condition we used face images where context dimensions were defined by social categories, similar to @canadas_social_2013. However, they found that social categorization occurred spontaneously and it was unclear what impact that would have on the current results. Therefore, we included another condition that used images of simple objects where context dimensions were defined by object features. Finally, @canadas_social_2013 also found that context effects generalized to novel face images. In our third condition we used non-repeating face exemplars to examine whether any task-relevance effects would also generalize in our design or if learning would be image-specific.

@@

```{r figure1, echo=FALSE, fig.width=6.7,fig.height=4.51, fig.cap='(ref:figure1)'}
knitr::include_graphics("images/Task.jpg")
```

@@

All participants were recruited from Amazon Mechanical Turk (AMT) and compensated $2.00 for participating. The amount compensated was calculated by estimating the maximum amount of time required to complete each experiment and multiplying by $6.00 per hour. For each experiment the number of HITs (Human intelligence tasks, an Amazon term for a work-unit) refers to the number of participants who initiated the study. Participants were included in the study if they completed all trials. We posted 150 HITs and 144 participants completed all trials.

@@

```{r analysis}
source("data-analysis/analysis.R")
```

```{r figure2, echo=FALSE,fig.width=6.875, fig.height= 4, fig.cap='(ref:figure2)'}
figure2
```

@@

@@

Similarly, the corresponding error analysis also resulted in a significant two-way interaction between the task-relevant context and unbiased-item congruency, `r ACC_ANOVA$Task_Relevant_Context_Congruency`, and a non-significant three-way interaction between task-relevant context, unbiased-item congruency, and context-type, `r ACC_ANOVA$Condition_Task_Relevant_Context_Congruency`. Therefore, the congruency effects in error rates were also smaller when the context dimension associated with high conflict was made task-relevant, corroborating the results of the reaction time analysis.

\begingroup
\setlength{\parindent}{0in}
\vspace{2.5mm}
\textbf{Table 1. \\ \textit{Reaction times and error rates from Experiment 1.}}

![ ](images/table1.pdf) 
\endgroup

@@

The aim of the current study was to determine whether task-relevance plays a role in the contextual recruitment of selective attention. Specifically, we tested whether manipulating the relative task-relevance of context cues could cause participants to ignore a previously learned context-association and apply a new association. We created a frequency unbiased context cue that shared features with two frequency biased contexts and used a feature-counting task to manipulate the task-relevance of context dimensions across two blocks of trials. Critically, halfway through the experiment participants received new instructions changing the task-relevant feature from one frequency biased cue to the other.

@@

In sum, our results provide new evidence that changes in task-relevance can update the contextual recruitment of selective attention in a CSPC flanker task. We demonstrated that making one context cue task-relevant produced a CSPC effect, even in the presence of a competing contextual cue. More important however, we found that changing the task-relevance of the contextual cues across blocks of trials was accompanied by predictable changes in the congruency effects. These effects were found to be generalizable across two different kinds of stimuli and occurred even when using non-repeating images, and implicate an important role for task-relevance in producing context-dependency and are consistent with memory-based accounts of CSPC phenomena.

@@

```{r tableLatex}
#if (knitr::opts_knit$get("rmarkdown.pandoc.to")== "latex"){
#source("data-analysis/table1.R")
#latexTable1
#}
```
